A Bayesian model of latitudinal temperature gradients applied to the Early Eocene

### 2023-02-12 manuscript/manuscript.Rmd
more discussion

The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC,  `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC  higher than the modern.

A model run excluding the ecological proxies results in a similar median latitudinal gradient, `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

@@

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).

These high-t proxies at high latitudes, combined with the overall high variability of the proxy record, is also the cause f...

@@

# Conclusions
In conclusion, the Bayesian hierarchical model presented in this study represents a significant improvement over existing methods for reconstructing past temperatures by incorporating both geochemical and ecological data. The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction. By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates. The hierarchical structure of the model accounts for the variation of temperature estimates at individual localities and uncertainty associated with the ecological temperature proxies. This improved method is expected to significantly alter the proxy-based temperature reconstructions of the past, particularly for intervals with small or biased samples of temperature proxies.
# Acknowledgements

>>>>

The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC,  `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC  higher than the modern. A model run excluding the ecological proxies results in a GMsST that is higher by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`) degC. The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

@@

[add table on lat gradients]

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC). The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies. Seasonality [@Keating-Bitoni] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

GMSsT

28 degC from bottom water temperatures

@@

# Conclusions

- focus on gradient rather than gmst - gmst can easily be calculated from gradient.

In conclusion, the Bayesian hierarchical model presented in this study represents a significant improvement over existing methods for reconstructing past temperatures by incorporating both geochemical and ecological data. The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction. By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates. The hierarchical structure of the model accounts for the variation of temperature estimates at individual localities and uncertainty associated with the ecological temperature proxies. This improved method is expected to significantly alter the proxy-based temperature reconstructions of the past, particularly for intervals with small or biased samples of temperature proxies.
# Acknowledgements
